---
title: "Bootstrap illustration"
author: "Benjamin Christoffersen"
date: "6 January 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::knit_hooks$set(default_opts = function(before, options, envir) {
    if (before){
      options(digist = 4)
    }
})
options(digist = 4)
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F, dpi = 36)
knitr::opts_knit$set(warning = F, message = F,  default_opts = T)
```

# Introduction

This vignette will breifly show how to bootstrap the estimate of a `ddhazard_fit`. You can get the version used to make this vignette by calling:`

```{r echo=FALSE}
tryCatch({
  current_sha <- paste0("@", httr::content(
    httr::GET("https://api.github.com/repos/boennecd/dynamichazard/git/refs/heads/master")
    )$object$sha)
}, error = function(...){ current_sha <<- "" })

stopifnot(length(current_sha) > 0 && class(current_sha) == "character")

current_version <- paste0("boennecd/dynamichazard", current_sha)
```

```{r}
warning("Building ddhazard")
current_version # the string you need to pass to devtools::install_github
```

```{r eval=FALSE}
devtools::install_github(current_version)
```

# PBC data set
We start by settings up the data set:
```{r}
# PBC data set from survival with time variying covariates
# Details of tmerge are not important in this scope. The code is included
# to make you able to reproduce the results
# See: https://cran.r-project.org/web/packages/survival/vignettes/timedep.pdf
library(survival)
temp <- subset(pbc, id <= 312, select=c(id, sex, time, status, edema, age))
pbc2 <- tmerge(temp, temp, id=id, death = event(time, status))
pbc2 <- tmerge(pbc2, pbcseq, id=id, albumin = tdc(day, albumin),
               protime = tdc(day, protime), bili = tdc(day, bili))
pbc2 <- pbc2[, c("id", "tstart", "tstop", "death", "sex", "edema", 
                 "age", "albumin", "protime", "bili")]
```

Next, we model:
```{r, fig.height = 4.5, mySettings=FALSE}
library(dynamichazard)
dd_fit <- ddhazard(Surv(tstart, tstop, death == 2) ~ age + edema +
                        log(albumin) + log(protime) + log(bili), pbc2,
                   id = pbc2$id, by = 100, max_T = 3600, 
                   Q_0 = diag(rep(10000, 6)), Q = diag(rep(0.001, 6)),
                   control = list(save_risk_set = T, save_data = T))

plot(dd_fit)
```
## Sampling individuals
Next, we bootstrap the estimates:
```{r}
set.seed(7451)
boot_out <- ddhazard_boot(
  dd_fit, 
  do_sample_weights = F,      # should re-sampeling be by weights or by
                              # sampling each individual discreetly
  do_stratify_with_event = F, # strafy on whether the individual is an event
                              # or not
  R = 500                     # Number of bootstrap samples
  )

# The list has the same structure and class the list returned by boot::boot
# Though, a few elements are added 
str(boot_out)
```
Above, bootsrap the model by sampeling the individuals discreetly. I.e. some individuals will have a weights $0,1,2,\dots$. We can plot 95% confidence bounds and illustrate the bootstrap estimates as follows:

```{r}
plot(dd_fit, ddhazard_boot = boot_out)
```

The completly black line is the estimates, the dashed lines are 95% confidence bounds taken at each point from the bootstrap estimates and the transparent black lines each represent a bootstrap estimate

Further, we can check how stratifying within those individual who are events and those who are not as follows:

```{r}
set.seed(8524)
boot_out_stratify_by_events <- ddhazard_boot(
  dd_fit, 
  do_sample_weights = F,
  do_stratify_with_event = T, # changed
  R = 500)

plot(dd_fit, ddhazard_boot = boot_out_stratify_by_events)
```

Lastly, we can also sample the weights. This is done as follows: within each strata (e.g. those who dies) let $q$ denote the number of individuals. Then we sample $q$ uniform variables $l_i~\text{Unif}(0,1)$ and normalize such that $\sum_{i=1}^q l_i = q$. If there is no stara then we use the entire sample of individuals. We can do this for our fit as follows:
```{r}
set.seed(401)
boot_out_by_weights <- ddhazard_boot(
  dd_fit, 
  do_sample_weights = T,      # changed      
  do_stratify_with_event = F, # set back to false
  R = 500)

plot(dd_fit, ddhazard_boot = boot_out_by_weights)
```

You can also provide your own strata argument by setting the `strata` argument in the call to `ddhazard_boot`. Note that this has to be on an individual level not observation level. Further, you can use the `unique_id` argument to match the individual entries with the entries in `strata`. As an example, we strafy by the age at the start of the study period with the code below:
```{r}
# First we show that all the individual has the same age for all periods
unique(tapply(pbc$age, pbc$id, function(x) length(unique(x))))

# Next, we find the age for each individual 
unique_id <- unique(pbc$id)
age <- sapply(unique_id, function(x) pbc$age[pbc$id == x][1])
summary(age)

# Make strata variable
is_less_than_50 <- age < 50

# Run bootstrap with strata variable
set.seed(101)
boot_out_with_strata <- ddhazard_boot(
  dd_fit,
  unique_id = unique_id, 
  strata = is_less_than_50,
  R = 500)

plot(dd_fit, ddhazard_boot = boot_out_with_strata)
```

The above is only for illustration. However, it may be usefull if you have e.g. catagorial variables in your model. Lastly, setting `do_stratify_with_event = T` will yield an interaction factor between the passed `strata` and whether or not the given individual do die
