---
title: "Bootstrap illustration"
author: "Benjamin Christoffersen"
date: "6 January 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::knit_hooks$set(default_opts = function(before, options, envir) {
    if (before){
      options(digist = 4)
    }
})
options(digist = 4)
knitr::opts_chunk$set(echo = TRUE, fig.height = 5, dpi = 36)
knitr::opts_knit$set(default_opts = T)
```

# Introduction

This vignette will breifly show how to bootstrap the estimate of a `ddhazard_fit`. It builds on the vignette 'ddhazard' and 'Comparing methods for time varying logistic models'. Thus, it is suggest first to read these. You can get the version used to make this vignette by calling:`

```{r echo=FALSE}
tryCatch({
  current_sha <- paste0("@", httr::content(
    httr::GET("https://api.github.com/repos/boennecd/dynamichazard/git/refs/heads/master")
    )$object$sha)
}, error = function(...){ current_sha <<- "" })

stopifnot(length(current_sha) > 0 && class(current_sha) == "character")

current_version <- paste0("boennecd/dynamichazard", current_sha)
```

```{r}
warning("Building ddhazard")
current_version # the string you need to pass to devtools::install_github
```

```{r eval=FALSE}
devtools::install_github(current_version)
```

# PBC data set
We start by settings up the data set:
```{r}
# PBC data set from survival with time variying covariates
# Details of tmerge are not important in this scope. The code is included
# to make you able to reproduce the results
# See: https://cran.r-project.org/web/packages/survival/vignettes/timedep.pdf
library(survival)
temp <- subset(pbc, id <= 312, select=c(id, sex, time, status, edema, age))
pbc2 <- tmerge(temp, temp, id=id, death = event(time, status))
pbc2 <- tmerge(pbc2, pbcseq, id=id, albumin = tdc(day, albumin),
               protime = tdc(day, protime), bili = tdc(day, bili))
pbc2 <- pbc2[, c("id", "tstart", "tstop", "death", "sex", "edema", 
                 "age", "albumin", "protime", "bili")]
```

Next, we fit the model as in the vignette 'Comparing methods for time varying logistic models':
```{r}
library(dynamichazard)
dd_fit <- ddhazard(Surv(tstart, tstop, death == 2) ~ age + edema +
                        log(albumin) + log(protime) + log(bili), pbc2,
                   id = pbc2$id, by = 100, max_T = 3600, 
                   Q_0 = diag(rep(10000, 6)), Q = diag(rep(0.001, 6)),
                   control = list(save_risk_set = T, save_data = T, eps = .1))

plot(dd_fit)
```

## Sampling individuals
We can bootstrap the estimates in the above model by :
```{r}
set.seed(7451)
R <- 10000
boot_out <- ddhazard_boot(
  dd_fit, 
  do_sample_weights = F,      # should re-sampeling be by weights or by
                              # sampling each individual discreetly
  do_stratify_with_event = F, # stratify on whether the individual is an event
                              # or not
  R = R                       # Number of bootstrap samples
  )

# The list has the same structure and class as the list returned by boot::boot
# Though, a few elements are added 
str(boot_out)
```
Above, we bootsrap the model by sampeling the individuals discreetly. I.e. individuals will have weights of $0,1,2,\dots$. We can plot 95% confidence bounds and illustrate the bootstrap estimates as follows:

```{r}
plot(dd_fit, ddhazard_boot = boot_out)
```

The completly black line is the estimates, the dashed lines are 2.5% and 97.5% quantiles of the bootstrap estimates taken at each point and the transparent black lines each represent a bootstrap estimate

We can check if it affects the results if we sample in strata between those individual who had an event and those who are not as follows:

```{r}
set.seed(8524)
boot_out_stratify_by_events <- ddhazard_boot(
  dd_fit, 
  do_sample_weights = F,
  do_stratify_with_event = T, # changed
  R = R)

plot(dd_fit, ddhazard_boot = boot_out_stratify_by_events)
```

## Sampling weights
We can also sample the weights. This is done as follows: within each stratum $j$ (e.g. those who have an event and those who do not) let $q_j$ denote the number of individuals. Then we sample $q_j$ uniform variables $l_i~\text{Unif}(0,1)$ for $i = 1,\dots,q_j$ and normalize such that $\sum_{i=1}^q l_i = q$. If there is no stara then we have single stratum with the entire sample of individuals. The code below will sample the weights as described above:
```{r}
set.seed(401)
boot_out_by_weights <- ddhazard_boot(
  dd_fit, 
  do_sample_weights = T,      # changed      
  do_stratify_with_event = F, # set back to false
  R = R)

plot(dd_fit, ddhazard_boot = boot_out_by_weights)
```

## Ohter strata
You can also provide your own strata by setting the `strata` argument in the call to `ddhazard_boot`. Note that this has to be on an individual level not observation level. Further, you can use the `unique_id` argument to match the individual entries with the entries in `strata`. As an example, we strafy by the age at the start of the study period with the code below:
```{r}
# First we show that all the individual has the same age for all periods
unique(tapply(pbc2$age, pbc2$id, function(x) length(unique(x))))

# Next, we find the age for each individual 
unique_id <- unique(pbc2$id)
age <- sapply(unique_id, function(x) pbc2$age[pbc$id == x][1])
summary(age)

# Make strata variable
is_less_than_50 <- age < 50

# Run bootstrap with strata variable
set.seed(101)
boot_out_with_strata <- ddhazard_boot(
  dd_fit,
  unique_id = unique_id, 
  strata = is_less_than_50,
  R = R)

plot(dd_fit, ddhazard_boot = boot_out_with_strata)
```

The above is only for code is only provide for illustrative purposes. There is no reason to stratify accross the age. However, it may be usefull if you have e.g. catagorial variables in your model. Lastly, setting `do_stratify_with_event = T` will yield an interaction factor between the passed `strata` and whether or not the given individual do die

## Fixed effects
Fixed effects wil also be sampled. They are added as the last entries of the element `t` returned by `ddhazard_boot`. As an example ...:

```{r}
dd_fit <- ddhazard(Surv(tstart, tstop, death == 2) ~ 
                     ddFixed(1) +ddFixed(log(protime)) # changed to fixed
                   + age + edema + log(albumin) +  + log(bili), pbc2,
                   id = pbc2$id, by = 100, max_T = 3600, 
                   Q_0 = diag(rep(10000, 4)), Q = diag(rep(0.001, 4)),
                   control = list(
                     save_risk_set = T, save_data = T, eps = .1,
                     fixed_terms_method = "E_step") # Fixed effects are 
                                                    # esimated in E-step
                   )

plot(dd_fit)
```

```{r}
# The fixed effects estimates
dd_fit$fixed_effects
```

```{r}
set.seed(9001)
boot_out <- ddhazard_boot(
  dd_fit, 
  do_sample_weights = T,      # sample using weights      
  do_stratify_with_event = F, # dont stratify by event
  R = R)

# Plot time varying effects
plot(dd_fit, ddhazard_boot = boot_out)
```

```{r, fig.height=3.5}
# Print some of bootstrap estimates of the fixed effects
fixed_effects_ests <- t(tail(t(boot_out$t), 2))
head(fixed_effects_ests)

# Make a boxplot of the fixed effects
par(mfcol = c(1, 2))
for(i in 1:ncol(fixed_effects_ests)){
  boxplot(fixed_effects_ests[, i], main = colnames(fixed_effects_ests)[i])
  
  # Add a horisontal line at the estimates with the whole sample
  abline(h = dd_fit$fixed_effects[i])
}
```
